Remote sensing riverine gas exchange and estimating carbon efflux using SWOT observations

### 2021-03-05 manuscript_vc/manuscript_v1.Rmd
manuscript v2

##Key Points (85 characters each)
- We introduce the BIKER algorithm to remotely sense riverine gas exchange using the SWOT satellite
- BIKER reproduces estimates of $CO_2$ fluxes when coupled with in situ gas concentration data
- Expected SWOT measurement errors exert a marginal influence on BIKER's estimates
- BIKER's accuracy is limited by process-level uncertainties and not by its algorithmic setup

@@

##1 Introduction
The inland water drainage network both transports and transforms various constiutents as they move from the landscape to the ocean [@colePlumbingGlobalCarbon2007a]. As a result, rivers and lakes are usually supersaturated with dissolved greenhouse gases (GHGs) and present a significant flux of these gases from water to air [@raymondGlobalCarbonDioxide2013a]. This flux $[M/L^2T]$ is calculated as equation 1 for the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$. Note that *k* is generally studied as normalized by a Schmidt number of 600 ($k_{600}$) to be gas and temperature agnostic [see @hallGasExchangeStreams2020 for a thorough review of gas exchange in rivers].

@@

It is infeasible to measure $k_{600}$ directly when working at the drainage-network-scale and so ‘upscaling functions’ are used that predict $k_{600}$ from readily available river geomorphology [e.g. @oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a]. Many upscaling functions are based on the basic theory that *k* is empirically correlated with the turbulent energy dissipation *e* at the air-water interface [@zappaEnvironmentalTurbulentMixing2007]. This is convenient for upscaling because in rivers and streams *e* can be equated to the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters: $eD \approx gSV$ where *g* is gravitational acceleration $[L/T^2]$, *S* is channel slope, and *V* is average flow velocity $[L/T]$ [@tsivoglouTracerMeasurementReaeration1976]. Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by the quality of their estimates of the velocity of the river channel. This is exasperated in ungauged basins where velocity models cannot be trained because no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

@@

McFLIs often employ Bayesian inference. Bayesian techniques have been previously used to estimate $k_{600}$ concurrently with stream metabolism from dissolved oxygen ($O_2$) datasets [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010]. While this approach requires detailed field data and is thus not applicable to $k_{600}$ estimation in ungauged rivers, it does suggest that Bayesian techniques could be useful for estimating $k_{600}$ when parameter equifinality is a problem (as it is with McFLIs).

In this context, we hypothesize that SWOT data will improve our understanding of riverine GHG fluxes in ungauged basins and/or at drainage-network-scales after techniques have been developed to ingest SWOT's novel measurements of water surface width and height. Therefore, we sought to develop a method to estimate $k_{600}$ using only SWOT measurements and no in situ information. We hypothesize that remote sensing $k_{600}$ is possible by coupling Bayesian remote sensing techniques (in a manner similar to McFLI) and SWOT data with hydraulic geometry and gas exchange theory. More specifically, this manuscript aims to answer two questions: <br>
	1) is ungauged RS of the gas exchange velocity possible using soon-to-be-available SWOT river measurements? <br>
	2) how might this algorithm influence upscaled estimates of riverine carbon efflux to the atmosphere?

@@

While equation 2 would be ideal for the goal of this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of *eD* to assign its differential model parameters. We do not have this a priori knowledge in an ungauged, RS-based setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes). Thus, we need to extend the [@ulsethDistinctAirWater2019a] model to assign the two model parameters using only RS-able hydraulics. To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes. We use river width to define river size and train different upscaling functions for different groups of data with similar river sizes [Figure 2, using the data from @ulsethDistinctAirWater2019a]. In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).

@@

###2.2 Developing the BIKER algorithm
With a validated upscaling model that requires no a priori knowledge of *eD* (results presented in section 3.1) we turn to developing the algorithm to remotely sense $k_{600}$. The approach used here is strongly informed by @hagemannBAMBayesianAMHGManning2017a’s algorithm for ungauged RSQ via Bayesian inference. Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian inference problem, which we largely follow here to conceptualize remote sensing $k_{600}$ as a Bayesian inference problem that can be solved using SWOT data.

SWOT will provide river width and height (and thus water surface slope $S_h$) at nodes/cross-sections along a set of river reaches. Thus, our ‘data’ for the Bayesian inference are SWOT measurements of river width and water surface slope while all other terms are treated as statistical parameters. We also obtain a ‘change in channel area’ from the river height and width and assuming a rectangular channel geometry.

In order to use Bayesian inference to solve for $k_{600}$, we need to derive a sampling model/likelihood function. Following @tsivoglouTracerMeasurementReaeration1976, $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for an algorithm ingesting SWOT data). Using this, as well as Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 3 as equation 4. In equation 4, we use the Manning’s formulation from Hagemann et al. (2017) which formulates depth via river width and area (in order to use SWOT-observables). $k_{600}$ is now defined as a function of $S_h$, channel roughness n $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ (discretized via the change in channel area *dA* and median channel area $A_0$). It is also defined by the upscaling parameters $a_i$ and $b_i$.

$$k_{600}=a_i(gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3})^{b_i} \mathbf{(4)}$$

Equation 4 is then reworked into a Bayesian sampling model with the ‘data’ on the left-hand side sampled from the unknown model parameters on the right-hand side (equation 5). Note that in the formal likelihood specification both sides of equation 5 are ln-transformed to produce the normal sampling model presented here, however they are written here without that transformation for succinctness. The $σ_{upscale}^2$ parameter refers to the uncertainty inherent in the upscaled $k_{600}$ estimates themselves (equation 4).

$$a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i} \sim normal(k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}, \sigma^2_{upscale})  \mathbf{(5)}$$

Equation five necessitates that Bayesian priors be assigned to model parameters *n*, $A_0$, and $k_{600}$. Bayesian priors formularize the a priori estimates (and uncertainties) for these parameters. Or more intuitively for these physcial quantities, they represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed. All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for $\lambda < X < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for parameter *X*. In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only. *n* and $A_0$ prior hyperparameters were assigned following the method developed by @brinkerhoffConstrainingRemoteRiver2020. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. With this method, they showed substantial improvement in the accuracy of their ungauged RSQ estimates with no changes to the algorithms or data used. We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the @ulsethDistinctAirWater2019a dataset to assign $\mu$ (equation 6). $σ^2$ was set to the standard error of equation 6: 1.023. λ and ε were set to -3 and 2.7, respectively.

@@

Finally, we estimate $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4. Specifically, we explore three uncertainty scenarios to assess the realtive inlfuences of the different sources of uncertainity. Uncertainty in equation 4 can stem from three sources: 1) error baked into the upscaling model ($a_i$ and $b_i$ parameter uncertainty), 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error. Our interest is using BIKER for satellite remote sensing and so we consider measurement errors from SWOT-derived water surface slopes. To do this, we run 15,000 unique MC simulations on 5,000 sets of field measurements of river channel hydraulics [@brinkerhoffReconcilingAtaStationAtManyStations2019] under three different uncertainty scenarios. Each MC simulation is itself 10,000 runs, sampling from the normal distributions for $a_i$ and $b_i$ (uncertainty scenario 1), $a_i$, $b_i$, and ln-transformed velocity (uncertainty scenario 2), and $a_i$, $b_i$, ln-transformed velocity, and water surface slope (uncertainty scenario 3). Estimate uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error to be 0.25. The slope-error model that we used is explained in detail in Section 2.3 and relies on a recent analysis of SWOT slope measurement errors [@durandHowWillRadar2020a]. This results in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term for each  scenario is extracted and used for $σ_{upscale}^2$.

@@

Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on ‘SWOT-like data’. There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces measurement errors to these simulated rivers. Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible. These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as ‘RS observations’ and are used as the sole inputs to the BIKER algorithm. Simulated rivers mimic perfect measurement conditions (i.e. no measurement errors). Here, we use the 22 rivers archived by @frassonCompilationHydraulicModels2019a and used by @rodriguezObservingRiversVarying2020 to develop reach-averaged Saint Vernant equations for the SWOT mission. We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model (**is that sufficient to say?**) and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately aligns the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4). All told, this yielded 503 sets of SWOT observations to validate BIKER with.

@@

To explore the effect of SWOT measurement errors on BIKER, the SWOT simulator mentioned above is the ideal tool. However, it is cumbersome to run on many rivers and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT water surface slope/height errors [@durandHowWillRadar2020a]. They developed a first-principles model for layover error from SWOT observations and, using global hydrography, quantified SWOT measurement error from both error tolerances and radar layover for 220,924 river reaches. The 68th percentiles of these reach-scale errors were 1.7cm/km and 10.4cm for water surface slope and height, respectively. To apply this model to our 22 rivers, we generate new river heights and water surface slopes by sampling every observation from normal distributions following equations 7 and 8 for node *i* and timestep *t*. Because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors, again following [@durandHowWillRadar2020a] and our rectangular channel assumptions: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$. Note that equation 7 is also used to define slope measurement error in the MC analysis (section 2.2).

@@

$$H_{err_{i,t}} \sim normal(H_{obs_{i,t}}, 0.104^{2}) \mathbf{(8)}$$

Regardless of the validation data used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides. Further, we are principally interested in our ability to reproduce upscaled $k_{600}$ from RS observations and are less concerend with the actual accuracy of the upscaling model itself [which can be validated using existing datasets like previously done by @raymondScalingGasTransfer2012a and @ulsethDistinctAirWater2019a]. Therefore, we take the best performing upscaling model from sections 2.1 and 3.1 and use that to calculate the ‘observed’ $k_{600}$ that BIKER is validated against. This is done using equation 9, where $V_{obs} [L/T]$ is observed river discharge divided by observed channel area. With this setup, we can directly explore BIKER's ability to infer observed *V* and $k_{600}$ from river width and height alone. It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning’s equation (and not $a_i$ and $b_i$). **MIGHT NEED TO ACCOUNT FOR SLOPE UNCERTAINITY TOO? I can probably justify not doing this though**. Thus, $\sigma_{upscale}^2$ is set to 0.25 for this validation. However in practice, it should reflect the total uncertainty calculated in section 2.2 from all three sources.

@@

Validation metrics take two forms (and are detailed in Table 1). To validate across all rivers and timesteps, we used the coefficient of determination $r^2$ and the root mean square error (RMSE). Four normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value (respectively) to compare across rivers. rBIAS is a measure of prediction bias that is normalized by the mean observed value to compare across rivers. KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than 0 is often interpreted as a useful prediction in ungauged settings, and a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean value [@knobenTechnicalNoteInherent2019a].

*Table 1: Validation metrics used in this study, where r is the correlation coefficient, Nt is number of observations and i is the specific observation. σ refers to the variance of the sample and μ refers to the mean of the sample.*

@@

###2.4 Upscaling to CO2 evasion and bulk carbon efflux
It is one thing to to predict $k_{600}$, but researchers are usually more interested in the gas fluxes from rivers and ultimately the bulk carbon efflux from river to atmosphere. Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2 [M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux $[M/T]$ (via $CO_2$ evasion) to the method used to estimate reach averaged flow velocity.

First, we calculated $FCO_2$. To do this, we paired the modeled $k_{CO2}$ values (obtained from the $k_{600}$ estimates) with field-measurements of water-side $CO_2$ concentrations and water temperatures. 26 bi-weekly samples were made by @beaulieuControlsGasTransfer2012a- Figure S1) at one location in the Ohio River for one calender year from 2008-2009. We paired these 26 values with each 11-day-repeat set of SWOT observations by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and this was deemed acceptable). Not all of the SWOT rivers have observations for a full year, and when simulation dates were not available they were assumed to start on January 1st. Note that this $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes multiple sections of the Ohio River). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable. Atmospheric $CO_2$ was assumed 390 uatm. The Schmidt number, used to obtain $k_{CO2}$ from $k_{600}$, was calculated following @raymondScalingGasTransfer2012a and @wanninkhofRelationshipWindSpeed1992a.

Then, we estimated bulk carbon efflux using four models for average flow velocity:  BIKER and three gauge-based HG models previously used for $CO_2$ upscaling [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a]. See Table S1 for their definitions. It is worth stressing that the @lauerwaldSpatialPatternsCO22015 model was developed as one of two components of the @raymondGlobalCarbonDioxide2013a model, and so was actually developed by the latter’s authors. The names used here refer to the specific implementation used in each study. For our study, all HG methods use the in-situ discharge record while BIKER does not. This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method (BIKER) versus gauged methods (all three HG models). Finally, we express the bulk carbon efflux as the average mass flow rate of carbon (via $CO_2$ evasion) per year from the 22 rivers after accounting for total river surface area.

>>>>

##Highlights (3-5 points, 85 characters each w/ spaces)
- The BIKER algorithm can remotely sense riverine gas exchange using the SWOT satellite
- BIKER predicts $CO_2$ fluxes when coupled with in situ gas concentration data
- Predicted $CO_2$ fluxes are similar/marginally worse than when using in-situ methods
- Expected SWOT measurement errors exert a marginal influence on BIKER's estimates
- Accuracy is limited by process-level uncertainties and not by the algorithmic setup

##Keywords
gas exchange, SWOT, remote sensing, carbon, ungauged basin, biogeochemistry

@@

##1 Introduction
The inland water drainage network both transports and transforms various constiutents as they move from the landscape to the ocean [@colePlumbingGlobalCarbon2007a]. As a result some of these processes, rivers and lakes are usually supersaturated with dissolved greenhouse gases (GHGs) and present a significant flux of these gases from water to air [@raymondGlobalCarbonDioxide2013a]. This flux $[M/L^2T]$ is calculated as equation 1 for the gas concentration gradient between the water $C_{water} [M/L^3]$ and the air $C_{air} [M/L^3]$ and the gas exchange velocity $k [L/T]$. Note that *k* is generally studied as normalized by a Schmidt number of 600 ($k_{600}$) to be gas and temperature agnostic [see @hallGasExchangeStreams2020 for a thorough review of gas exchange in rivers].

@@

It is infeasible to measure $k_{600}$ directly when working at the drainage-network-scale and so ‘upscaling functions’ are used that predict $k_{600}$ from readily available river geomorphology [e.g. @oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a]. Many upscaling functions are based on the basic theory that *k* is empirically correlated with the turbulent energy dissipation *e* at the air-water interface [@zappaEnvironmentalTurbulentMixing2007]. This is convenient for upscaling because in rivers and streams *e* can be equated to the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters: $eD \approx gSV$ where *g* is gravitational acceleration $[L/T^2]$, *S* is channel slope, and *V* is average flow velocity $[L/T]$ [@tsivoglouTracerMeasurementReaeration1976]. Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by the quality of their estimates of the reach-averaged flow velocity. This is exasperated in ungauged basins where velocity models cannot be trained because no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

@@

McFLIs often employ Bayesian inference. Bayesian techniques have been previously used to estimate $k_{600}$ concurrently with stream metabolism from dissolved oxygen ($O_2$) datasets [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010]. While this approach requires detailed field data and is thus not applicable to $k_{600}$ estimation in ungauged rivers, it does suggest that Bayesian techniques could be useful for remotely sensing $k_{600}$ when parameter equifinality is a problem as it is with McFLIs [@andreadisConstrainingAssimilationSWOT2020d]. Equinality issues refer to under-constrained systems that can have multiple solutions to the same discharge problem [@garamboisInferenceEffectiveRiver2015].

In this context, we hypothesize that SWOT data will improve our understanding of riverine GHG fluxes in ungauged basins and/or at drainage-network-scales we develop techniques to ingest SWOT's novel measurements of water surface width and height. Therefore, we sought to develop a method to estimate $k_{600}$ using only SWOT measurements and no in situ information. We hypothesize that remote sensing $k_{600}$ is possible by coupling Bayesian remote sensing techniques (in a manner similar to McFLI) and SWOT data with hydraulic geometry and gas exchange theory. More specifically, this manuscript aims to answer two questions:

- Is ungauged RS of the gas exchange velocity possible using soon-to-be-available SWOT river measurements?
- How might this algorithm influence upscaled estimates of riverine carbon efflux to the atmosphere?

@@

While equation 2 would be ideal for the goal of this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of *eD* to assign its differential model parameters. We do not have this a priori knowledge in an ungauged, RS-oriented setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes). Thus, we need to extend the @ulsethDistinctAirWater2019a model to assign the two model parameters using only RS-able hydraulics. To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes. We use river width to define river size and train different upscaling functions for different groups of data with similar river sizes [Figure 2, using the data from @ulsethDistinctAirWater2019a]. In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).

@@

###2.2 Developing the BIKER algorithm
With a validated upscaling model that requires no a priori knowledge of *eD* (results presented in section 3.1) we turn to developing the algorithm to remotely sense $k_{600}$. The approach used here is strongly informed by @hagemannBAMBayesianAMHGManning2017a’s algorithm for ungauged RSQ via Bayesian inference. Consult that paper for the logic behind conceptualizing the equation for discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data.

SWOT will provide river width and height (and thus water surface slope $S_h$) at nodes/cross-sections along a set of river reaches. Thus, our ‘data’ for the Bayesian inference are SWOT measurements of river width and water surface slope while all other terms are treated as statistical parameters that cannot be directly observed from SWOT. We also obtain a ‘change in channel area’ *dA* from the river height and width and assuming a rectangular channel geometry.

In order to use Bayesian inference to solve for $k_{600}$, we need a sampling model/likelihood function that estimates the unobservable statisitcal parameters. Following @tsivoglouTracerMeasurementReaeration1976, $eD \approx gS_hV$ for unsteady flow conditions (channel slope $S_0 \neq S_h$- see text S1 for this model derivation and why assuming non-uniform flow is convenient for an algorithm ingesting SWOT data). Using this, as well as Manning’s equation to relate average flow velocity to channel geometry, we rewrite equation 3 as equation 4. In equation 4, we use the Manning’s formulation from Hagemann et al. (2017) which formulates depth via river width and area (in order to use SWOT-observables). $k_{600}$ is now defined as a function of $S_h$, channel roughness *n* $[T/L^{\frac{1}{3}}]$, channel width *W* [L], and channel area $A_0+dA [L^2]$ (discretized via median channel area $A_0$ and the previously introduced *dA*). It is also defined by the upscaling parameters $a_i$ and $b_i$.

$$k_{600}=a_i\left[gS_h^{\frac{3}{2}}\frac{1}{n}(A_0+dA)^\frac{2}{3}W^\frac{-2}{3}\right]^{b_i} \mathbf{(4)}$$

Equation 4 is then reworked into a Bayesian sampling model to have all of the ‘data’ on the left-hand side sampled from all the unknown model parameters on the right-hand side (equation 5, where *x* is the SWOT observables and $\Theta$ are a set of statistical parameters). Note that in the formal likelihood specification both sides of equation 5 are ln-transformed to produce the normal sampling model presented here, however they are written in equation 5 without that transformation for succinctness. The $σ_{upscale}^2$ parameter refers to the uncertainty inherent in the upscaling model itself (equation 4). This will be explained in detail below.

$$f(x|\Theta) = \left[a_i^{-1}(gS_h^\frac{-3}{2}W^\frac{2}{3})^{b_i}\right] \sim normal\left(\left[k_{600}^{-1}\frac{1}{n}(A_0+dA)^\frac{2}{3})^{b_i}\right], \sigma^2_{upscale}\right)  \mathbf{(5)}$$

Equation five necessitates that Bayesian priors be assigned to model parameters *n*, $A_0$, and $k_{600}$. Bayesian priors formularize the a priori estimates (and uncertainties) for the non-remotely-sensed parameters. Or more intuitively for these physcial quantities, they represent our ‘prior river knowledge’ of what *n*, $A_0$, and $k_{600}$ probably are for some river given that they cannot be remotely sensed. All priors are formalized within the model as truncated normal distributions of the ln-transformed terms such that $ln(X) \sim N(\mu, \sigma^2)$ for $\lambda < X < \epsilon$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\epsilon$) and lower bounds ($\lambda$) for parameter *X*. In order to avoid relying on in situ information, we assign prior hyperparameters using SWOT data only. *n* and $A_0$ prior hyperparameters were assigned following the method developed by @brinkerhoffConstrainingRemoteRiver2020. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. With this method, they showed substantial improvement in the accuracy of their ungauged RSQ estimates with no changes to the algorithms or data used. We assigned the $k_{600}$ prior hyperparameters using a simple slope regression model trained on the @ulsethDistinctAirWater2019a dataset to assign $\mu$ (equation 6). $σ^2$ was set to the standard error of equation 6: 1.023. λ and ε were set to -3 and 2.7, respectively.

@@

Finally, we estimate $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4. Specifically, we explore three uncertainty scenarios to assess the realtive influences of different sources of uncertainity. Uncertainty in equation 4 can stem from 1) error baked into the upscaling model ($a_i$ and $b_i$ parameter uncertainty), 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error. Our interest is using BIKER for satellite remote sensing and so we consider measurement errors from SWOT-derived water surface slopes. To do this, we run 15,000 unique MC simulations on 5,000 sets of field measurements of river channel hydraulics from @brinkerhoffReconcilingAtaStationAtManyStations2019 under three different uncertainty scenarios. Each MC simulation is itself 10,000 runs, sampling from the normal distributions for $a_i$ and $b_i$ (uncertainty scenario 1), $a_i$, $b_i$, and ln-transformed velocity (uncertainty scenario 2), and $a_i$, $b_i$, ln-transformed velocity, and water surface slope (uncertainty scenario 3). Estimate uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error to be 0.25. The slope-error model that we used is explained in detail in Section 2.3 and relies on a recent analysis of SWOT water surface slope measurement errors [@durandHowWillRadar2020a]. This results in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term for any scenario is extracted and used for $σ_{upscale}^2$. **a table might help here to layout the three tests**

@@

Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on ‘SWOT-like data’. There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces measurement errors to these simulated rivers. Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible. These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as ‘RS observations’ and are used as the sole inputs to the BIKER algorithm. Simulated rivers mimic perfect measurement conditions (i.e. no measurement errors). Here, we use the 22 rivers archived by @frassonCompilationHydraulicModels2019a and used by @rodriguezObservingRiversVarying2020 to develop reach-averaged Saint Vernant equations for the SWOT mission. We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model **is that sufficient to say?** and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately aligns the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4). All told, this yielded 503 sets of SWOT observations to validate BIKER with.

@@

To explore the effect of SWOT measurement errors on BIKER, the SWOT simulator mentioned above is the ideal tool. However, it is cumbersome to run on many rivers **is this sufficient?** and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT water surface slope/height errors [@durandHowWillRadar2020a]. They developed a first-principles model for layover error from SWOT observations and, using global hydrography, quantified SWOT measurement error from both error tolerances and radar layover for 220,924 river reaches. The 68th percentiles of these reach-scale errors were 1.7cm/km and 10.4cm for water surface slope and height, respectively. To apply this model to our 22 rivers, we generate new SWOT observations by sampling every observation from normal distributions following equations 7 and 8 for node *i* and timestep *t*. Note that because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors again following @durandHowWillRadar2020a and our rectangular channel assumptions: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$. Note that equation 7 is also used to define slope measurement error in the MC analysis (section 2.2).

@@

$$dA_{err_{i,t}} \sim normal(dA_{obs_{i,t}}, \left[W_{obs_{i,t}}*\sqrt{2}*0.104\right]^{2}) \mathbf{(8)}$$

Regardless of the validation data used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides. Further, we are principally interested in our ability to reproduce upscaled $k_{600}$ from RS observations and are less concerend with the actual accuracy of the upscaling model itself [which can be validated using existing datasets like previously done by @raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a]. Therefore, we take the best performing upscaling model from sections 2.1 and 3.1 and use that to calculate the ‘observed’ $k_{600}$ that BIKER is validated against. This is done using equation 9, where $V_{obs} [L/T]$ is observed river discharge divided by observed channel area. With this setup, we can directly explore BIKER's ability to infer observed *V* and $k_{600}$ from river width and height alone. It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning’s equation (and not $a_i$ and $b_i$). Thus, $\sigma_{upscale}^2$ is set to 0.25 for this validation. However in practice, it should reflect the total uncertainty calculated in section 2.2 from all three sources.

@@

Validation metrics take two forms (and are detailed in Table 1). To validate across all rivers and timesteps, we used the coefficient of determination $r^2$ and the root mean square error RMSE. Four normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value (respectively) to compare across rivers. rBIAS is a measure of prediction bias that is normalized by the mean observed value to compare across rivers. KGE is a standard metric used in streamflow prediction with an intuitive basis: a value greater than 0 is often interpreted as a useful prediction in ungauged settings, and a value greater than -0.41 indicates a model outperforms a uniform prediction of the mean observed value [@knobenTechnicalNoteInherent2019a].

*Table 1: Validation metrics used in this study, where r is the correlation coefficient, Nt is number of observations and i is the specific observation. σ refers to the variance of the sample and μ refers to the mean of the sample. As is standard, a carrot accent indicates the predicted value.*

@@

###2.4 Upscaling to CO2 evasion and bulk carbon efflux
It is one thing to to predict $k_{600}$, but researchers are usually more interested in the gas fluxes from rivers and ultimately the bulk carbon efflux from river to atmosphere. Therefore, we explored 1) the ability of BIKER to reproduce hypothetical carbon dioxide evasion fluxes ($FCO_2$ $[M/L^2T]$) from these 22 rivers, and 2) the sensitivity of the estimated bulk carbon efflux $[M/T]$ via $CO_2$ evasion to the method used to estimate reach averaged flow velocity.

First, we calculated $FCO_2$. To do this, we paired the modeled $k_{CO2}$ values (obtained from $k_{600}$) with field-measurements of water-side $CO_2$ concentrations and water temperatures. 26 bi-weekly samples were made by @beaulieuControlsGasTransfer2012a- Figure S1) at one location in the Ohio River for one calender year from 2008-2009. We paired these 26 values with each 11-day-repeat set of SWOT observations by date, ignoring the timesteps beyond 26 (only ~15% of the SWOT observations were ignored here and we deemed this acceptable). Not all of the SWOT rivers have observations for a full year, and when simulation dates were not available they were assumed to start on January 1st. Note that this $CO_2$ data is for the Ohio River only but was applied to all 22 rivers (which includes multiple sections of the Ohio River- @frassonCompilationHydraulicModels2019a). Because we are exclusively interested in the relative differences between $FCO_2$ estimates and not the raw fluxes themselves, any $CO_2$ data representative of SWOT-observable rivers was deemed acceptable. Atmospheric $CO_2$ was assumed 390 uatm. The Schmidt number, used to backcalculate $k_{CO2}$ from $k_{600}$, was calculated following @raymondScalingGasTransfer2012a and @wanninkhofRelationshipWindSpeed1992a.

Then, we estimated bulk carbon efflux using four models for average flow velocity:  BIKER and three gauge-based HG models previously used for $CO_2$ upscaling [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a; @raymondScalingGasTransfer2012a]. See Table S1 for their definitions. It is worth stressing that the @lauerwaldSpatialPatternsCO22015 model was developed as one of two components of the @raymondGlobalCarbonDioxide2013a model, and so was actually developed by the latter’s authors. The names used here refer to the specific implementation used in each study. For this study, all HG methods use the in-situ discharge record while BIKER does not. This allows us to explore the relative differences in bulk carbon efflux estimates if implementing a wholly ungauged method (BIKER) versus gauged methods (all three HG models). Finally, we express the bulk carbon efflux as the average mass flow rate of carbon (via $CO_2$ evasion) per year from the 22 rivers after accounting for total river surface area.
